---
title: "Getting Started with vota"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with vota}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**vota** is an R package for simulating Spanish electoral outcomes. Starting from national survey data (transfer matrices), it projects vote intentions to provincial seat assignments using the D'Hondt method -- the same allocation system used in Spanish elections.

This vignette walks you through the basic setup and a minimal simulation using the package's included datasets.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("hmeleiro/vota")
```

```{r}
library(vota)
```

## Included Datasets

The package includes real data from the Spanish general elections of July 23, 2023. These datasets serve as both examples and ready-to-use inputs:

### Transfer Matrix (`mt`)

The transfer matrix captures how voters from the last election intend to vote now. Each row represents a combination of past party recall (`recuerdo`) and current vote intention (`idv`), along with the percentage transferring between them.

```{r}
data(mt)
head(mt)
```

### Previous Election Results (`votos_23J`)

Official vote totals from the 23J election, used as the base for demographic adjustments and voter redistribution.

```{r}
data(votos_23J)
votos_23J
```

### Provincial Voting Patterns (`patrones_23J`)

Historical voting distributions by province, used to project national estimates to the provincial level.

```{r}
data(patrones_23J)
head(patrones_23J)
```

### Seats per Province (`n_seats`)

The number of Congressional seats allocated to each of Spain's 52 electoral districts.

```{r}
data(n_seats)
head(n_seats)
```

### Manual Adjustments (`retoques`)

Optional expert adjustments -- add or subtract votes from specific parties based on external knowledge.

```{r}
data(retoques)
retoques
```

### Small Parties (`small_parties`)

Vote estimates for parties too small to model in the transfer matrix but relevant for the overall picture.

```{r}
data(small_parties)
small_parties
```

## Setting Up a Project

The easiest way to start a simulation is to scaffold a project directory:

```{r eval=FALSE}
setup_electoral_project("my_simulation_2024")
```

This creates:

- `input/input.xlsx` -- An Excel template pre-filled with example data
- `output/` -- Directory for simulation results
- `scripts/main.R` -- A starter script

Alternatively, generate just the Excel template:

```{r eval=FALSE}
create_input_template("input/input.xlsx")
```

## Running a Simulation

Once you have your input data in an Excel file, run the full pipeline with `run_vota()`:

```{r eval=FALSE}
results <- run_vota(
  input_path = "input/input.xlsx",
  output_file = "output/results.rds",
  uncertainty_method = "mcmc",
  nsims = 100,
  factor_correccion_abstencion = 3,
  factor_correccion_jovenes = 2.5,
  factor_correccion_otbl = 2.5,
  tiempo_entre_elecciones = 0.1,
  tau = 300,
  umbral = 0.03,
  seed = 42,
  verbose = TRUE
)
```

The function returns an `electo_fit` object containing:

- **`estimacion`** -- National estimates with confidence intervals (vote % and seats)
- **`estimacion_sims`** -- All simulation results
- **`estimacion_provincias_sims`** -- Provincial-level simulation results
- **`dhondt_output`** -- Detailed D'Hondt allocation
- **`mt_sims_pct`** / **`mt_sims_electores`** -- Transfer matrix simulations
- **`participacion_media`** -- Estimated mean turnout
- **`metadata`** -- Execution parameters

## Inspecting Results

```{r eval=FALSE}
# Quick overview
print(results)

# Detailed summary with win probabilities
summary(results)

# Access the national estimates table
results$estimacion
```

## Visualizing Results

The `electo_fit` object has a dedicated `plot()` method with four visualization types:

```{r eval=FALSE}
# National vote shares with confidence intervals
plot(results, "nacional")

# Distribution of seats across simulations (by party)
plot(results, "seats_dist")

# Seats per province for a specific party
plot(results, "provincia", partido = "PP")

# Seat fragility: margin of last D'Hondt quotient
plot(results, "dhondt_margin")
```

## Next Steps

- **[The VOTA Methodology](methodology.html)** -- Understand the five corrections and the statistical model
- **[Step-by-step Tutorial](tutorial.html)** -- Walk through each function in the pipeline using the internal datasets
