---
title: "Step-by-step Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Step-by-step Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial walks through the vota pipeline step by step, using only the datasets included in the package. Instead of running `run_vota()` as a black box, we'll call each function individually to understand what happens at each stage.

```{r}
library(vota)
library(dplyr)
library(tidyr)
```

## Step 1: Explore the Internal Data

### The Transfer Matrix

The `mt` dataset is an example transfer matrix in wide format. Each row is a vote intention party (`idv`), and each column is a previous vote recall party. The last column (`idv == "N"`) provides the sample size for each column (vote recall).

```{r}
data(mt)
mt
```

The columns represent parties from the *previous* election (recuerdo): PSOE, PP, Vox, Sumar, OTBL (other/blank), and ABNL (abstention/null). The `idv` column shows who these voters now intend to vote for. For example, the row where `idv = "PP"` shows how many voters of each previous party now intend to vote PP.

### Previous Election Results

`votos_23J` contains the official total votes per party from the 23J 2023 election:

```{r}
data(votos_23J)
votos_23J
```

These totals are used to convert transfer percentages into absolute vote counts.

### Provincial Patterns

`patrones_23J` contains each party's historical vote share by province (as a proportion between 0 and 1):

```{r}
data(patrones_23J)
head(patrones_23J, 12)

# How many provinces and parties?
cat("Provinces:", length(unique(patrones_23J$codigo_provincia)), "\n")
cat("Parties:", length(unique(patrones_23J$partido)), "\n")
```

### Seats per Province

```{r}
data(n_seats)
head(n_seats)
cat("Total seats:", sum(n_seats$n_diputados), "\n")
```

### Optional: Adjustments and Small Parties

```{r}
data(retoques)
retoques

data(small_parties)
small_parties
```

## Step 2: Prepare the Transfer Matrix

The `simulate_mt()` function needs the transfer matrix in long format with columns `recuerdo`, `idv`, `n`, and `pct_original`. Let's prepare it from the wide `mt` dataset.

```{r}
# Convert wide mt to long format
party_cols <- setdiff(names(mt), c("idv", "n"))

mt_long_raw <- 
  mt %>%
  pivot_longer(
    cols = party_cols,
    names_to = "recuerdo",
    values_to = "pct_original"
  ) 

mt_long <- 
  mt_long_raw %>% 
  filter(idv != "N") %>% 
  left_join(
    mt_long_raw %>% 
      filter(idv == "N") %>% 
      select(recuerdo, n = pct_original), 
    by = join_by(recuerdo)
  ) %>%
  select(recuerdo, idv, n, pct_original)

head(mt_long)
```

The `pct_original` column contains the transfer percentage for each recuerdo-idv pair, and `n` contains the number of survey respondents in that cell.

## Step 3: Simulate Transfer Matrices

Using `simulate_mt()`, we generate multiple simulated transfer matrices via multinomial resampling. Each simulation draws from the sample distribution, creating natural variability.

```{r}
mt_sims <- simulate_mt(mt_long, nsims = 5, seed = 42)

# sim = 0 is the original (unperturbed) matrix
head(mt_sims)
cat("Number of simulations (including original):", length(unique(mt_sims$sim)), "\n")
```

Each simulation (`sim = 1, 2, ...`) is a plausible transfer matrix. `sim = 0` is the original observed matrix.

## Step 4: Run the VOTA Algorithm

The `vota()` function takes a single transfer matrix (in long format with `recuerdo`, `idv`, `transfer` columns) and returns national vote estimates. Let's run it on the original matrix (`sim = 0`):

```{r}
# Get the original (sim=0) transfer matrix
mt_original <- mt_sims %>% filter(sim == 0)

result <- vota(
  mt_simplificada = mt_original,
  tiempo_entre_elecciones = 0.1,
  factor_correccion_abstencion = 3,
  factor_correccion_jovenes = 2.5,
  factor_correccion_otbl = 2.5,
  retoques = retoques,
  small_parties = small_parties,
  votos_ant = votos_23J
)

# National vote estimates
result$estimacion %>% 
  arrange(desc(votos))
```

The output is a data frame with estimated votes per party (`idv`). These are absolute vote counts.

We can compute vote percentages:

```{r}
estimacion <- result$estimacion %>%
  mutate(pct = votos / sum(votos) * 100) %>%
  arrange(desc(pct))

estimacion
```

## Step 5: Run VOTA Across All Simulations

To quantify uncertainty, we run `vota()` on each simulated transfer matrix:

```{r}
all_estimates <- lapply(unique(mt_sims$sim), function(s) {
  mt_s <- mt_sims %>% filter(sim == s)
  res <- vota(
    mt_simplificada = mt_s,
    tiempo_entre_elecciones = 0.1,
    factor_correccion_abstencion = 3,
    factor_correccion_jovenes = 2.5,
    factor_correccion_otbl = 2.5,
    retoques = retoques,
    small_parties = small_parties,
    votos_ant = votos_23J
  )
  res$estimacion %>% mutate(sim = s)
})

all_estimates <- bind_rows(all_estimates)

# Compute percentages per simulation
all_estimates <- all_estimates %>%
  group_by(sim) %>%
  mutate(pct = votos / sum(votos) * 100) %>%
  ungroup()

head(all_estimates, n = 20)
```

Now we can see how estimates vary across simulations:

```{r}
summary_stats <- all_estimates %>%
  group_by(idv) %>%
  summarise(
    pct_median = median(pct),
    pct_lwr = quantile(pct, 0.05),
    pct_upr = quantile(pct, 0.95),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_median))

summary_stats
```

## Step 6: Provincial Projection

National estimates need to be distributed across Spain's 52 provinces using historical voting patterns. The `simulate_prov_votes()` function does this using a Dirichlet distribution:

```{r}
# Use the point estimate (sim=0)
estimacion_nacional <- result$estimacion

# Filter patterns to only include parties in our estimation
parties_in_estimation <- estimacion_nacional$idv
patrones_filtrados <- patrones_23J %>%
  pivot_longer(
    all_of(parties_in_estimation), 
    names_to = "idv", 
    values_to = "patron"
  ) %>%
  mutate(patron = patron / 100) %>%
  filter(patron > 0)

# Simulate provincial vote distribution
prov_votes <- simulate_prov_votes(
  patrones = patrones_filtrados,
  estimacion = estimacion_nacional,
  method = "dirichlet",
  tau = 300,
  seed = 42
)

# Result is a matrix: provinces (rows) x parties (columns)
dim(prov_votes)
head(prov_votes[, 1:4])
```

Each cell contains the simulated number of votes for that party in that province. The `tau` parameter controls how closely the provincial distribution follows the historical pattern (higher = less variability).

## Step 7: D'Hondt Seat Allocation

Finally, we allocate seats using the D'Hondt method. First, we prepare the data in the format expected by `fast_dhondt()`:

```{r}
# Convert provincial votes matrix to long format
prov_df <- as.data.frame(prov_votes) %>%
  mutate(codigo_provincia = rownames(prov_votes)) %>%
  pivot_longer(
    cols = -codigo_provincia,
    names_to = "partido",
    values_to = "votos_prov"
  ) %>%
  # Add seat counts

  left_join(n_seats, by = "codigo_provincia") %>%
  # Add simulation identifier
  mutate(sim = 0L) %>%
  # Calculate vote percentage per province
  group_by(sim, codigo_provincia) %>%
  mutate(
    votos_validos = sum(votos_prov),
    pct_sobre_validos = votos_prov / votos_validos
  ) %>%
  ungroup()

head(prov_df)
```

Apply the electoral threshold (3%) and run D'Hondt:

```{r}
# Filter parties that pass the threshold
prov_above_threshold <- prov_df %>%
  filter(!partido %in% c("OTBL", "ABNL"), pct_sobre_validos >= 0.03)

# Run D'Hondt allocation
dhondt_result <- fast_dhondt(
  data = prov_above_threshold,
  cod_prov = codigo_provincia,
  sim = sim,
  partido = partido,
  votos_prov = votos_prov,
  nseats = n_diputados
)

head(dhondt_result)
```

Aggregate to get total seats per party:

```{r}
seats_by_party <- dhondt_result %>%
  filter(tipo == "Asignado") %>%
  count(partido, name = "seats") %>%
  arrange(desc(seats))

seats_by_party
cat("Total seats assigned:", sum(seats_by_party$seats), "\n")
```

## Step 8: The Full Pipeline with `run_vota()`

In practice, you would use `run_vota()` which orchestrates all of the above steps automatically. It reads from an Excel file, so let's create one using the template function:

```{r eval=FALSE}
# Create a temporary project
tmp_dir <- tempdir()
project_dir <- file.path(tmp_dir, "tutorial_project")
setup_electoral_project(project_dir)

# Run the full pipeline
results <- run_vota(
  input_path = file.path(project_dir, "input", "input.xlsx"),
  output_file = file.path(project_dir, "output", "results.rds"),
  uncertainty_method = "mcmc",
  nsims = 50,
  factor_correccion_abstencion = 3,
  factor_correccion_jovenes = 2.5,
  factor_correccion_otbl = 2.5,
  tiempo_entre_elecciones = 0.1,
  tau = 300,
  umbral = 0.03,
  seed = 42,
  verbose = TRUE
)

# View results
print(results)
summary(results)

# Visualize
plot(results, "nacional")
plot(results, "seats_dist")
```

## Summary

In this tutorial we walked through each step of the electoral simulation:

1. **Data exploration** -- Understanding the transfer matrix, previous results, and provincial patterns
2. **Transfer matrix simulation** -- `simulate_mt()` for multinomial uncertainty
3. **VOTA algorithm** -- `vota()` with five sequential corrections
4. **Provincial projection** -- `simulate_prov_votes()` with Dirichlet variability
5. **Seat allocation** -- `fast_dhondt()` for vectorized D'Hondt
6. **Full pipeline** -- `run_vota()` tying everything together

## Further Reading

- **[Getting Started](getting-started.html)** -- Quick setup guide
- **[The VOTA Methodology](methodology.html)** -- Statistical details of the algorithm
